%% Uniformly automatic classes of finite groups of bounded rank-width
%% ------------------------------------------------------------------
%% REWRITE (v2 of the draft, 2026-07): ring-first presentation over
%% R = Z/p^d, one master theorem (tree, distributed centers) with the
%% string version as a corollary, warm-ups with mini-proofs, and the
%% product-of-constructions section absorbed from scratch-chainring.tex.
%% The previous draft (field-only, three separate theorems) is preserved
%% in git history as v1; comments below cite it as "v1" for reuse.
%%
%% STATUS OF SECTIONS
%%   1 Introduction ............... WRITTEN
%%   2 Preliminaries .............. WRITTEN (chain ring, forms over R, cut-rank)
%%   3 Warm-ups ................... WRITTEN (two mini-proofs + ring bridge)
%%   4 Linear layouts ............. WRITTEN (row-module interfaces over R)
%%   5 Master theorem ............. WRITTEN (restriction calculus, pairing
%%                                  tables, claim modules; string + fixed-center
%%                                  corollaries)
%%   6 Wealth of subclasses ....... WRITTEN (interface, products, tables,
%%                                  widths, collapse, generic, dividing line)
%%   7 Consequences ............... WRITTEN
%%   8 Implementation ............. WRITTEN (numbers marked for refresh from
%%                                  the extended benchmark before submission)
%%   9 Open problems .............. WRITTEN
%%
%% TODO before submission/arXiv: affiliation, ORCID, funding, ccsdesc check,
%% verify all bibliography details against DBLP, event macros.
%% CAMERA-READY (once accepted): remove the `anonymous' class option below.
\documentclass[a4paper,UKenglish,autoref,anonymous]{lipics-v2021}

\usepackage{microtype}
\usepackage{booktabs}

\bibliographystyle{plainurl}

\title{Taming the Wild Stratum: Uniformly Automatic Class-2 Groups of Bounded Rank-Width}
\titlerunning{Groups of Bounded Rank-Width}

\author{Faried Abu Zaid}{Independent Researcher, Germany}{fariedaz@gmail.com}{}{}

\authorrunning{F. Abu Zaid}

\Copyright{Faried Abu Zaid}

\ccsdesc[500]{Theory of computation~Finite Model Theory}
\ccsdesc[300]{Theory of computation~Regular languages}
\ccsdesc[300]{Computing methodologies~Symbolic and algebraic algorithms}

\keywords{automatic structures, advice automata, finite $p$-groups, rank-width, chain rings, group isomorphism, first-order model checking}

\supplementdetails[subcategory={Source Code}]{Software}{https://github.com/fariedabuzaid/AutStr}

%% TODO: replace by the actual event once the venue is fixed.
\EventEditors{}
\EventNoEds{0}
\EventLongTitle{}
\EventShortTitle{Preprint}
\EventAcronym{}
\EventYear{2026}
\EventDate{}
\EventLocation{}
\EventLogo{}
\SeriesVolume{0}
\ArticleNo{0}

%% ------------------------------------------------------------------
\newcommand{\Fp}{\mathbb{F}_p}
\newcommand{\Zp}{\mathbb{Z}_p}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\R}{R}                                   % the chain ring Z/p^d
\newcommand{\rk}{\operatorname{rk}}
\newcommand{\rs}{\operatorname{rowsp}}
\newcommand{\cs}{\operatorname{colsp}}
\newcommand{\mrk}{\rho}                              % module cut-rank
\newcommand{\lcr}{\operatorname{lcr}}
\newcommand{\tcr}{\operatorname{tcr}}
\newcommand{\adv}{\alpha}
\newcommand{\conv}{\otimes}
\newcommand{\lin}[1]{\mathcal{C}^{\mathrm{lin}}_{#1}}
\newcommand{\tre}[1]{\mathcal{C}^{\mathrm{tree}}_{#1}}
\newcommand{\site}[1]{\mathcal{C}^{\mathrm{site}}_{#1}}
%% The software library is authored by us; under the anonymous class option we
%% hide its (googlable) name and suppress the self-citation whose bib entry
%% carries our name and repository URL. Dropping the option restores both.
\ifx\authoranonymous\relax
  \newcommand{\AutStr}{\textcolor{red}{[name omitted for review]}}
  \newcommand{\AutStrCite}{}
\else
  \newcommand{\AutStr}{\textsf{AutStr}}
  \newcommand{\AutStrCite}{~\cite{AutStr}}
\fi

\begin{document}

\maketitle

\begin{abstract}
Finite nilpotent groups of class~$2$ dominate the enumeration of finite
groups; their isomorphism problem is complete for the tensor-isomorphism
class, and by Mekler's construction the variety interprets all graphs. We
identify a structural parameter under which this wild stratum becomes
uniformly tame for automata-based methods: the \emph{rank-width of the
commutation data}, measured over the chain ring $\R = \ZZ/p^d$ as the free
rank of the crossing modules. Our main theorem is a single general
statement: for fixed $p$, $d$ and $r$, the class-$2$ groups with center of
exponent $p^d$ whose commutation tensor---central generators distributed
arbitrarily over a layout tree---admits a layout in which every subtree
cut's six crossing flattenings have module rank at most $r$ form a
uniformly tree-automatic class, and on chain layouts a uniformly
word-automatic one. The automaton state \emph{is} the cut interface, so
the parameter is not merely sufficient but literally the machine.
Two ring phenomena separate $\ZZ/p^d$ from the field and shape the
construction: a sibling merge, though determined by the carried interface
values, need not be a bilinear form \emph{in} them---the letters carry
bounded pairing tables instead of matrices---and the expectations that
distributed centers impose on their environment are managed as elements
of bounded \emph{claim modules}, with a unique-extension property
supplying determinism where coordinates fail to exist.
Consequences include linear-time first-order model checking on every
member and decidability of the uniform theory of each class. The class is
closed under central and direct products, and many natural families of
class-$2$ groups---none of which had a uniform automatic presentation
before---arise as low-width layouts: the extraspecial $p$-groups (whose
direct limit $E_p$ is the one previously known word-automatic class-$2$
group), tree-indexed laminar groups with growing center, groups whose
commutation graph has bounded pathwidth, treewidth or vertex cover, and
the clique, on which no two generators commute. Generic forms have width
$\Omega(n)$ under every layout, and Mekler's graph-interpreting groups,
ultraspecial field-multiplication tensors and the relatively free
class-$2$ group lie outside. All constructions are implemented and machine-verified in the
\AutStr{} library; the layout compilers derive every advice constant by
solving a linear system whose solvability is one of our factorisation
lemmas, so each successful compilation certifies the lemmas on that
instance. We leave open a converse in the spirit of Courcelle--Oum and
the complexity of isometry for matrix spaces of bounded rank-width.
\end{abstract}

\section{Introduction}
\label{sec:intro}

Two research programmes meet in this paper.

\subparagraph*{Algorithmic group theory.}
Among finite groups, the nilpotent groups of class~$2$ and small exponent
are simultaneously the most numerous and the least classifiable. The
Higman--Sims bounds show that the number of groups of order $p^m$ is
$p^{\frac{2}{27}m^3 + O(m^{8/3})}$, and the lower bound is already attained
by class-$2$ groups of exponent $p$~\cite{Higman1960,Sims1965,Pyber1993}; it
is a long-standing folklore conjecture that almost all finite groups are of
this kind. By the Baer correspondence~\cite{Baer1938}, such a group is
determined by an alternating bilinear map $B \colon \Fp^n \times \Fp^n \to
\Fp^k$, and its isomorphism problem is exactly the isometry problem for
alternating matrix spaces, which is complete for the tensor-isomorphism
class~\cite{GrochowQiao2021}. Model theory tells the same story from another
side: by Mekler's construction, this variety of groups interprets all graphs
and is as wild as first-order model theory permits~\cite{Mekler1981}. In
short, class-$2$ groups of small exponent are the dark matter of finite
group theory: they form the bulk of the universe and resist classification.

\subparagraph*{Algorithmic model theory.}
Automatic structures~\cite{KhoussainovNerode1995,BlumensathGraedel2000,Graedel2020}
present a structure by finite automata: elements are words, and each
relation is recognised by a synchronous multi-tape automaton. Automaticity
buys decidable first-order theories and linear-time query evaluation. For a
\emph{class} of finite structures the right notion is \emph{uniform}
automaticity with advice~\cite{AbuZaidGraedelReinhardt2017,AbuZaid2018}: one
tuple of automata presents every member of the class, and an additional
advice string, read synchronously with the element encodings, selects the
member. Uniformly automatic classes admit algorithmic meta-theorems in the
style of Courcelle's theorem~\cite{Courcelle1990}: every first-order
property can be compiled, once and for all, into an automaton that runs in
linear time on each member's advice. The published stock of uniformly
automatic classes consists of all finite abelian groups, all finite boolean
algebras, and the graphs of bounded tree-depth~\cite{AbuZaid2018}. On the
group-theoretic side, automatic presentations are known to hug
commutativity closely: finitely generated word-automatic groups are
virtually abelian~\cite{OliverThomas2005}, and the recent positive result
of Nies and Stephan concerns a \emph{single} infinite structure---the
infinite extraspecial $p$-group $E_p$ is word-automatic---while the
relatively free class-$2$ exponent-$p$ group on infinitely many generators
is not~\cite{NiesStephan2023}. No uniformly automatic class of
non-abelian groups has appeared in print; everything of that kind in this
paper is proved from the published baseline just described.

\subparagraph*{From one group to a class, and from a matching to a tree.}
Our starting point is the observation that the positive Nies--Stephan
result is, secretly, about a family. $E_p$ is the direct limit of the
finite extraspecial groups of order $p^{1+2m}$, and advice turns the limit
into the family: a single automaton with \emph{two} $\Fp$-registers, reading
one advice letter per generator, multiplies in every finite extraspecial
group at once (\autoref{prop:warmA}). The construction is a five-line
streaming argument---one register buffers the digit just read, the other
carries a running deficit---and it provokes the question this paper
answers: \emph{what do the registers measure?}

A first hint that the answer is not ``the number of active generators''
comes from growing the center. Place one central generator at every leaf
of a binary shape tree, a generator pair at every inner node, and let each
pair's commutator hit exactly the central generators below it. The center
of this \emph{tree-indexed} group grows linearly with the group, its
commutation data is laminar rather than a matching---and a single
$\Fp$-register still suffices for a bottom-up automaton, provided sibling
subtrees are forced to agree on the deficit they owe to the leaves beneath
them (\autoref{prop:warmB}).

The answer, and the subject of this paper, is that in both warm-ups the
commutation data crossing any cut of the layout has \emph{rank one}. The
registers never store digits; they store the image of the digits read so
far under a rank factorisation of the \emph{crossing block} at the current
cut. What a streaming evaluator of the group law must carry across a cut
is exactly this block, and if it factors through rank $r$, then $r$
residues suffice, whatever the number of active generators. Bounding the
cut-rank over the cuts of a linear or tree layout is bounding the
\emph{rank-width of the commutation form}---the same step that graph
theory takes from treewidth to rank-width and
clique-width~\cite{OumSeymour2006,CourcelleOum2007}, taken here inside
group theory.

\subparagraph*{This paper.}
We prove one general theorem and specialise it. Fix a prime $p$, a depth
$d \ge 1$ and a width $r$. A \emph{site tree} is a binary tree whose nodes
are generators ($x$-sites) or central generators ($z$-sites); a
\emph{cocycle tensor} over the chain ring $\R = \ZZ/p^d$ assigns to each
generator pair and each central target a coefficient in $\R$, presenting a
class-$2$ group whose center has exponent $p^d$. For every subtree cut,
the tensor entries crossing the cut sort into \emph{six flattenings}---
upward digit functionals on each side, resolved pair-sums, inward claims,
and two mixed kinds---and the \emph{width} of the layout is the maximal
\emph{module rank} of these flattenings: the minimal number of
generators of the spanned module, which over a field is the ordinary
rank.
Our master theorem (\autoref{thm:master}) states that for fixed $p, d, r$
the pairs (site tree, tensor of width $\le r$) form a uniformly
tree-automatic class, with a multiplication automaton of $p^{O(rd)}$
states whose state \emph{is} the cut interface; on chain-shaped layouts
the presentation is word-automatic (\autoref{cor:string}), and with the
central generators on a chain above the layout it specialises to clean
fixed-center theorems whose linear case (\autoref{thm:word}) we prove
first as a fully worked simple instance.

Two algebraic obstacles separate the master theorem from its field,
fixed-center shadow, and their resolutions are the technical heart of the
paper. First, the \emph{ring}: over a field, the interface carried across
a cut can be normalised to a basis, and the bilinear pairing of two
sibling interfaces at a merge is again a matrix over the field. Over
$\ZZ/p^d$ both steps fail, already at width one and depth two
(\autoref{ex:ringfail}): normalised interfaces (bases of pure closures)
are non-unique and do not restrict into one another from one cut to the
next, while the honest interface---a generating set of the crossing
\emph{row module}---may carry valuation, and then the merge value, though
completely determined by the two interface values, is not a bilinear form
in them: over $\ZZ/4$ the pairing $2ab$ of the interface values $w = 2a$
and $v = 2b$ is the well-defined function $2 \cdot (w/2)(v/2)$, and no
$q$ satisfies $2ab = q\,w v$. The construction therefore carries row-module
generators, whose restriction calculus is exact over any ring
(\autoref{lem:restrict}), and lets merge letters carry bounded
\emph{pairing tables} rather than matrices (\autoref{lem:tables}).
Second, the \emph{distributed center}: commutator targets inside a
subtree can be owed contributions by pairs resolved \emph{outside} it, an
inward flow that no bottom-up pass receives. The automaton meets it with
a deterministic \emph{claim-and-verify} protocol: the residuals of the
central checks inside a subtree must lie in a bounded \emph{claim module},
the automaton carries them as an element of that module, and determinism
comes from a unique-extension property (\autoref{lem:claims})---over a
field this is the familiar ``commit to coordinates in a full-column-rank
basis'', but over the ring coordinates need not exist and the module
formulation is the correct one. Both mechanisms are invisible at $d = 1$,
where every nonzero scalar is a unit; both are forced over $\ZZ/4$
already.

\subparagraph*{Contributions.}
\begin{itemize}
\item \emph{The master theorem} (\autoref{thm:master}): for all fixed $p,
  d, r$, the class-$2$ groups with exponent-$p^d$ center and a site-tree
  layout of module width $\le r$ are uniformly tree-automatic; chain
  layouts give word-automaticity (\autoref{cor:string}). The fixed-center
  specialisations recover, over $\R$, the natural word and tree classes of
  bounded (linear) rank-width (\autoref{thm:word}, \autoref{cor:tree}).
\item \emph{The ring calculus of interfaces}: module cut-rank as the
  correct width; the exact restriction calculus of row-module interfaces
  (\autoref{lem:restrict}); the failure of matrix merges over $\ZZ/p^d$
  and their replacement by bounded pairing tables (\autoref{lem:tables});
  and claim modules with unique-extension determinism for distributed
  centers (\autoref{lem:claims}).
\item \emph{The wealth of the class} (\autoref{sec:span}): closure under
  central and direct products (width is the maximum, not the sum, over
  blocks); uniform versions of extraspecial groups, Heisenberg groups,
  tree-indexed laminar and point-target families, shared-center graph
  groups of bounded graph rank-width, and commutation graphs of bounded
  pathwidth, treewidth or vertex cover---one decidable family with a
  modular advice. On the excluded side: generic forms have width
  $\Omega(n)$ under every layout, and Mekler's construction, ultraspecial
  groups and the relatively free group mark the boundary.
\item \emph{Meta-theorems} (\autoref{sec:consequences}): compiled
  first-order model checking in time $O(\log|G|)$ per member, and
  decidability of the uniform theory of each class, including extensions
  by cardinality and Ramsey quantifiers.
\item \emph{Machine verification and feasibility}
  (\autoref{sec:implementation}): all constructions, including the general
  protocol over the ring, are implemented in the open-source \AutStr{}
  library\AutStrCite{}; the compilers are self-certifying instances of the
  factorisation lemmas, and a benchmark maps which classes and queries
  compile in practice and how far implicit evaluation reaches beyond.
\end{itemize}

\subparagraph*{What the parameter costs, honestly.}
For the group theorist we are explicit about the price of tameness. At
$k = 1$ and odd $p$ the symplectic normal form collapses the fixed-center
class, up to isomorphism, to central products of Heisenberg groups with
abelian factors; bounded vertex cover collapses to a bounded core times an
elementary abelian factor (\autoref{prop:collapse}). The
isomorphism-theoretic content therefore begins at $k \ge 2$---where
normal forms are unavailable and a presentation that works \emph{without}
normalisation earns its keep---and, already at width one, with the
distributed centers of the master theorem, whose laminar and point-target
families have commutator subgroups of rank $\Theta(n)$. Conversely a
generic form has cut-rank $\Omega(n)$ under \emph{every} layout
(\autoref{prop:generic}): the class is a thin, structured slice of the
wild stratum, as any decidable slice must be.

\subparagraph*{Organisation.}
\autoref{sec:prelim} fixes notation and develops the little chain-ring
algebra we need (valuation, minimal generating sets, module cut-rank).
\autoref{sec:warmup} proves the two warm-ups and isolates, informally,
exactly what must generalise. \autoref{sec:word} proves the fixed-center
linear theorem over $\R$ as a fully worked simple case.
\autoref{sec:master} defines site trees, tensors and the six flattenings,
states the master theorem, and proves it; the string corollary and the
fixed-center corollaries close the section. \autoref{sec:span} maps the
class: corners, closure properties, classical width measures, collapses,
generic wideness, and the dividing line. \autoref{sec:consequences}
records the meta-theorems, \autoref{sec:implementation} the
implementation, machine verification and feasibility study, and
\autoref{sec:open} the open problems.

\section{Preliminaries}
\label{sec:prelim}

\subparagraph*{Words, trees, convolution.}
$\Sigma$ denotes a finite alphabet and $\Box \notin \Sigma$ a padding
symbol. The \emph{convolution} $w_1 \conv \dots \conv w_m$ of words is the
word over $(\Sigma \cup \{\Box\})^m$ obtained by aligning the $w_i$
position-wise and padding the shorter ones. A \emph{tree} is a finite,
binary, rooted, labelled tree; the convolution of trees overlays them and
pads missing positions. A (bottom-up deterministic) tree automaton
processes a tree from the leaves to the root; a run assigns a state to
every node as a function of the node's label and the states of its
children.

\subparagraph*{Uniformly automatic classes.}
We follow~\cite{AbuZaidGraedelReinhardt2017}. A \emph{uniformly automatic
presentation} of a class $\mathcal{C}$ of $\tau$-structures consists of a
regular advice set $A \subseteq \Sigma^*$ (or a regular set of advice trees)
and automata $\mathfrak{U}, (\mathfrak{A}_R)_{R \in \tau}$ such that for
every $\adv \in A$ the structure
\[
  \mathfrak{S}_\adv \;=\; \bigl( \{ u : \mathfrak{U} \text{ accepts } \adv
  \conv u \},\; \bigl( \{ \bar u : \mathfrak{A}_R \text{ accepts } \adv \conv
  \bar u \} \bigr)_{R \in \tau} \bigr)
\]
is well-defined, and $\mathcal{C} = \{ \mathfrak{S}_\adv : \adv \in A \}$ up
to isomorphism. The same structure may be presented by several advices. The
fundamental theorem of automatic structures relativises: for every
first-order formula $\varphi(\bar x)$ there is an automaton recognising
$\{ \adv \conv \bar u : \mathfrak{S}_\adv \models \varphi(\bar u) \}$,
effectively~\cite{AbuZaidGraedelReinhardt2017}. We present groups in the
signature $\{M, \mathrm{eq}\}$ with $M$ the graph of multiplication.

\subparagraph*{The chain ring and module rank.}
Throughout, $\R = \ZZ/p^d$ for a prime $p$ and a fixed depth $d \ge 1$; at
$d = 1$, $\R = \Fp$. $\R$ is a finite chain ring: a local principal ideal
ring whose ideals form the chain $\R \supsetneq p\R \supsetneq \dots
\supsetneq p^d\R = 0$. Every element is $u\,p^s$ with $u$ a unit and $0
\le s \le d$ its \emph{valuation} $v(\cdot)$; the units are exactly the
valuation-$0$ elements. We write $\rs(V)$ for the row module of a matrix
$V$ over $\R$ (the $\R$-span of its rows) and $\cs(V)$ for its column
module. For matrices $V, W$ with equally many columns, $W = TV$ is
solvable for a matrix $T$ iff every row of $W$ lies in $\rs(V)$; we use
this silently throughout.

\begin{lemma}[minimal generating sets]
\label{lem:smith}
Every submodule $M \subseteq \R^{m}$ has a generating set
$p^{e_1}u_1, \dots, p^{e_\mrk} u_\mrk$ in which $u_1, \dots, u_\mrk$
extend to a basis of $\R^m$ and $0 \le e_i < d$, where
$\mrk = \mrk(M) := \dim_{\Fp}\bigl(M/pM\bigr)$; and no generating set of
$M$ has fewer than $\mrk(M)$ elements.
\end{lemma}

\begin{proof}
Smith normal form over the principal ideal ring $\R$: $M =
\rs\bigl(\operatorname{diag}(p^{e_1}, \dots)\,U\bigr)$ for a unimodular
$U$, with the rows whose exponent satisfies $e_i \ge d$ vanishing and
dropped; the surviving rows of $U$ are the $u_i$. Their number is
$\dim_{\Fp} M/pM$, and Nakayama's lemma gives the lower bound for
arbitrary generating sets.
\end{proof}

We call $\mrk(M)$ the \emph{module rank} of $M$. At $d = 1$ it is the
dimension of a subspace; the point of the definition for $d > 1$ is that a
valuation-carrying generator still contributes a full unit of rank: over
$\ZZ/4$ the module generated by $(2, 0)$ has module rank $1$, although
its reduction mod $2$ is zero. \autoref{sec:warmup} shows that this is
exactly the right accounting for a streaming evaluator.

\subparagraph*{Forms and groups over $\R$.}
Fix $k \ge 1$. A \emph{form} of dimension $n$ over $\R$ is a family $B =
(B_{ji})_{1 \le i < j \le n}$ with $B_{ji} \in \R^k$; we write $\hat
B_{\{i,j\}} := B_{ji}$ for the unoriented label and treat absent labels
as $0$. Define the $\R$-bilinear map $C \colon \R^n \times \R^n \to \R^k$
by
\begin{equation}
  C(a, a') \;=\; \sum_{1 \le i < j \le n} B_{ji}\, a_j\, a'_i .
  \label{eq:cocycle}
\end{equation}

\begin{lemma}[normal forms]
\label{lem:cocycle}
The set $\R^k \times \R^n$ with the operation
\[
  (b, a) \cdot (b', a') \;=\; \bigl(b + b' + C(a, a'),\; a + a'\bigr)
\]
is a group $G(B)$ of order $p^{d(n+k)}$ and nilpotency class $\le 2$; the
elements $(e_l, 0)$ are central, $[\,(0, e_j),\, (0, e_i)\,] = (\hat
B_{\{i,j\}}, 0)$ for $i \ne j$, and every element has a unique normal
form $x_1^{a_1} \cdots x_n^{a_n}\, y_1^{b_1} \cdots y_k^{b_k}$ over the
generators $x_i = (0, e_i)$, $y_l = (e_l, 0)$.
\end{lemma}

\begin{proof}
Bilinearity of $C$ gives the cocycle identity $C(a, a') + C(a + a', a'')
= C(a, a'') + C(a', a'') + C(a, a')$, hence associativity; the inverse of
$(b, a)$ is $(-b + C(a, a), -a)$. Since $B_{ii}$ is not among the labels,
$C(e_i, e_i) = 0$, so $x_i$ has order $p^d$ exactly (also for $p = 2$);
commutators are central, so the class is at most $2$. The normal form is
immediate from the displayed law.
\end{proof}

Both coordinates range over $\R$, and this is forced: in a group of class
$\le 2$, $[x, y]^p = [x^p, y]$, so if every generator had order $p$ the
commutator subgroup would have exponent $p$ as well---a center of
exponent $p^d$ requires a quotient of exponent $p^d$. For odd $p$ and
$d = 1$, $G(B)$ has exponent $p$; for $p = 2$ central elements of higher
order may occur even at $d = 1$: with $n = 3$, $k = 1$ and all three
labels $1$, the element $z = x_1 x_2 x_3$ is central with $z^2 = y$, and
$G(B)$ is the $1$-qubit Pauli group. By the Baer
correspondence~\cite{Baer1938}, isomorphism of the groups $G(B)$ (for
odd $p$ and $d = 1$, modulo the radical) is isometry of the associated
alternating matrix spaces; all parameters introduced next are, up to the
choice of layout, isoclinism invariants~\cite{Hall1940}.

\subparagraph*{Cut-rank and layouts.}
For $S \subseteq [n]$ the \emph{crossing block} of $B$ at $S$ is the
matrix
\[
  M_S \in \R^{\,\bigl(([n] \setminus S) \times [k]\bigr) \times S},
  \qquad
  (M_S)_{(j,l),\,i} \;=\; \bigl(\hat B_{\{i,j\}}\bigr)_l ,
\]
and the \emph{module cut-rank} of $S$ is $\rho_B(S) :=
\mrk\bigl(\rs(M_S)\bigr)$. A \emph{linear layout} of $B$ is an ordering
of $[n]$; renaming generators, we always take it to be the natural
order, and define its width as $\max_{1 \le t < n} \rho_B([t])$. A
\emph{tree layout} is a binary rooted tree $T$ together with a bijection
of $[n]$ onto the nodes of $T$; renaming, we take the bijection to be
the post-order enumeration (left subtree, right subtree, node), so that
every subtree hosts an interval $[\mathrm{lo}_v, v]$. Its width is
$\max_v \rho_B(S_v)$ over the subtree generator sets $S_v$. We write
$\lcr(B)$ and $\tcr(B)$ for the minimal width over all linear (tree)
layouts; since a chain is a tree, $\tcr \le \lcr$. At $d = 1$ and $k =
1$ with $0/1$ labels, $M_S$ is the bipartite adjacency matrix between
$S$ and its complement in the commutation graph, and $\tcr$ coincides
with $\Fp$ rank-width (\autoref{sec:span} makes the interface exact).
For fixed $p, d, k, r$ let $\lin{p,k,r,d}$ denote the class of pairs
(form over $\R$, linear layout of width $\le r$), each presenting
$G(B)$, and $\tre{p,k,r,d}$ the tree analogue.

\section{Warm-up: two automata to grow}
\label{sec:warmup}

Neither result in this section is deep, and neither is in the literature:
the published automaticity results closest to them are about single
infinite structures~\cite{NiesStephan2023}. We prove them in full because
they are the intuition for everything that follows, and because the
master theorem will re-derive them as corner cases---the demonstration
that the general machine degenerates correctly.

\subparagraph*{A: the uniform extraspecial class.}
Fix a prime $p$. For $m \ge 1$ let $X_m$ be the \emph{matching form}: $n =
2m$ generators, $k = 1$, and $B_{2t,2t-1} = 1$ for $t \in [m]$, all other
labels zero. The presented group $G(X_m)$ of order $p^{1+2m}$ is, for odd
$p$, the extraspecial group of exponent $p$; for $p = 2$, the central
product of $m$ copies of $D_8$. The infinite extraspecial group $E_p$ of
Nies--Stephan is the direct limit of this family.

\begin{proposition}[the finite extraspecial family, uniformly]
\label{prop:warmA}
The class $\{G(X_m) : m \ge 1\}$ is uniformly word-automatic, with advice
$\mathsf{c}\,\sigma^{2m}$ for the $m$-th member and a multiplication
automaton with $2p^2 + 2$ states.
\end{proposition}

\begin{proof}
Encode $(b, a) \in \Fp \times \Fp^{2m}$ as the word $b\, a_1 \cdots
a_{2m}$. By the cocycle normal form (\autoref{sec:prelim}), $z = x \cdot
y$ holds iff $a_{z,i} = a_{x,i} + a_{y,i}$ position-wise and $b_z = b_x +
b_y + \sum_{t=1}^{m} a_{x,2t}\, a_{y,2t-1}$. The automaton keeps a deficit
$\delta \in \Fp$ and a buffer $w \in \Fp$. On the marker $\mathsf{c}$ it
sets $\delta := b_z - b_x - b_y$. On each $\sigma$ it checks the digit
sum; at odd positions ($2t{-}1$) it stores $w := a_{y,2t-1}$, at even
positions ($2t$) it updates $\delta := \delta - a_{x,2t} \cdot w$. It
accepts at the end iff $\delta = 0$. Parity is one bit of state; the
state count is $2 \cdot p^2$ live states plus an initial and a sink
state. Universe and equality automata check length parity and letter
ranges only.
\end{proof}

The construction generalises without comment to any \emph{matching}
form; what matters below is its shape: the buffer $w$ is a
\emph{one-dimensional interface}---the only information about the digits
read so far that the future will consume.

\subparagraph*{B: the laminar tree class.}
Now let the center grow. For a binary shape tree $T$, let $G_T$ be
generated by a pair $x_w, y_w$ for every \emph{inner} node $w$ of $T$ and
a central $z_v$ for every \emph{leaf} $v$, of exponent $p$, subject to
\[
  [\,x_w,\, y_w\,] \;=\; \prod_{\text{leaves } v \text{ below } w} z_v,
\]
all other pairs of generators commuting. The commutator subgroup is the
full center $\Fp^{\text{leaves}}$: it grows linearly with the group, and
the commutation targets form the laminar family of the tree. A spine
recovers $G(X_m)$-like groups with a single leaf; bushy trees do not
embed in any fixed-center family.

\begin{proposition}[the laminar tree family, uniformly]
\label{prop:warmB}
The class $\{G_T : T \text{ a binary shape tree}\}$ is uniformly
tree-automatic, with the shape tree itself as advice and a multiplication
automaton with $p + 1$ states.
\end{proposition}

\begin{proof}
Encode an element by labelling inner nodes $w$ with the digit pair
$(a_w, b_w)$ of $x_w, y_w$ and leaves $v$ with the central digit $c_v$;
the multiplication instance convolves three such labellings. Writing the
group law in normal form, $z = x \cdot y$ holds iff all digit sums match
position-wise and, for every leaf $v$,
\[
  c_{z,v} - c_{x,v} - c_{y,v} \;=\; \sum_{w \,\text{above}\, v}
  a_{x,w}\, b_{y,w}.
\]
The right-hand side depends only on the root path of $v$, and---this is
laminarity---it is the \emph{same} sum for every leaf below a given node.
So a bottom-up automaton can carry a single scalar: the deficit still
owed to \emph{all} leaves of the current subtree. At a leaf it starts as
$c_{z,v} - c_{x,v} - c_{y,v}$; at an inner node $w$ it must equal the
deficit of each child subtree---the automaton \emph{rejects if the two
children disagree}---and is decreased by $a_{x,w} b_{y,w}$, the
contribution of the pair at $w$ to everything below. The automaton
accepts at the root iff the owed deficit is $0$. States: the $p$ values
of the deficit plus a sink. Universe and equality automata check shapes
and letter ranges.
\end{proof}

\subparagraph*{What must generalise.}
The two warm-ups look different---one streams left to right with a fixed
center, the other climbs a tree whose center grows---but both automata
carry rank-one data across every cut, and each isolates one of the two
mechanisms of the master theorem.

In warm-up A the buffer is a \emph{functional export}: the cut after
position $2t-1$ is crossed by exactly one commutation pair, and $w$ is
the value of the (rank-one) crossing block applied to the digits inside
the cut. Replacing the matching by an arbitrary form, the crossing block
$M_S$ at a cut $S$ becomes an arbitrary matrix; an automaton with $r$
registers can carry $V_S \cdot (\text{digits in } S)$ for an $r$-row
matrix $V_S$ generating the row module of $M_S$, and the advice can spell
out how these interfaces transform from one cut to the next. That
programme, over the ring, is \autoref{sec:word}; on trees, where the two
interfaces of sibling subtrees must be paired against each other at
their merge, it is part of \autoref{sec:master}.

In warm-up B the deficit is a \emph{claim}: leaves inside the subtree are
owed contributions by pairs that live \emph{above}---outside---the
subtree, an inward flow that a bottom-up automaton cannot yet see. The
automaton copes by committing to what the outside must eventually
deliver: laminarity makes all leaves below a node interchangeable, so one
scalar describes the debt, and the sibling-agreement check is the
consistency of two commitments. In general the debts of the central
positions inside a cut form a vector constrained to a bounded
\emph{claim module}---the column module of a claim flattening---and the
automaton will carry that vector as an element of the module,
deterministically, rejecting when a residual falls outside it; every
merge will reconcile a child's claims against its sibling's exports.
That is the claim-and-verify protocol of \autoref{sec:master}.

\subparagraph*{Why the ring is not free.}
Both mechanisms survive the passage from center exponent $p$ to $p^d$---
from $\Fp$ to $\R = \ZZ/p^d$---but not by rewriting $\Fp$ as $\R$. Two
phenomena, both already visible at width one over $\ZZ/4$, have no field
analogue.

\begin{example}[merges are not matrices over $\ZZ/4$]
\label{ex:ringfail}
Consider a single commutation pair split between two sibling subtrees,
with label $2 \in \ZZ/4$, and suppose each subtree's interface is the
functional $w = 2a$ (respectively $v = 2b$) of the digit $a$
(respectively $b$) of its endpoint---the natural rank-one interface when
further pairs with label $2$ cross each cut. The merge must add the
contribution $2ab$. This value is completely determined by the two
interface values, since $2ab = 2\,(w/2)(v/2)$, where $w/2$ denotes the
well-defined residue $a \bmod 2$; but it is \emph{not} a bilinear form
in them: $q\,wv = 4qab = 0 \ne 2ab$ for every $q \in \R$. Pairing two
valuation-$1$ interfaces through a matrix only ever produces valuation
$\ge 2$; the valuation budget is spent. A merge over the ring must
therefore be allowed to apply an arbitrary function of the two register
values---a bounded \emph{pairing table}---rather than a matrix
(\autoref{lem:tables}).
\end{example}

\begin{example}[normalised interfaces do not restrict]
\label{ex:nonesting}
One might hope to restore matrix merges by normalising the interface: by
\autoref{lem:smith} every crossing module is contained in a free direct
summand of the same module rank (a \emph{pure closure}), and pairing
free interfaces is unproblematic. But pure closures are not unique---in
$(\ZZ/4)^3$, both $\R\,(1,0,1)$ and $\R\,(1,2,1)$ are minimal pure
overmodules of $\R\,(2,0,2)$---and they do not restrict into one
another. For the form $B_{31} = 2$, $B_{32} = 1$, $B_{42} = 2$ over
$\ZZ/4$, of width one, the crossing module at the cut $\{1,2,3\}$ is
$\R\,(0,2,0)$, with pure closure $\R\,(0,1,0)$; restricting its
generator to the first two coordinates gives $(0,1)$, which lies in no
pure closure of the crossing module $\R\,(2,1)$ at the cut
$\{1,2\}$---indeed $\R\,(2,1)$ is already pure, and $(0,1) \notin
\R\,(2,1)$. An automaton carrying the normalised functional $a_2 \bmod
4$ at the larger cut would need it delivered by an interface that can
only ever supply $2a_1 + a_2$; no transition matrix exists. The
row-module generator $(0,2,0)$, by contrast, restricts to $(0,2) = 2
\cdot (2,1)$, and the transition is the multiplication by $2$.
\end{example}

The two examples fix the design: the interfaces carried across cuts are
\emph{generating sets of the crossing row modules}, for which restriction
is exact over any ring (\autoref{lem:restrict}), and the merges consume
them through tables (\autoref{lem:tables}). The claims of warm-up~B meet
the same issue in a different guise---over the ring the residual vector
inside a subtree has no canonical coordinates---and are carried as
elements of a bounded \emph{claim module}, with determinism supplied by a
unique-extension property (\autoref{lem:claims}). With these
adjustments, everything in the warm-ups scales.

\section{Linear layouts over the chain ring}
\label{sec:word}

Throughout this section fix $p, d, k, r$ and a form $B$ over $\R$ with a
linear layout of width $\le r$; write $M_t := M_{[t]}$.

The multiplication automaton must verify $(b_z, a_z) = (b_x, a_x) \cdot
(b_y, a_y)$ while reading the three encodings synchronously, digit by
digit. The digit sums $a_{z,t} = a_{x,t} + a_{y,t}$ are position-wise;
the work is the correction $C(a_x, a_y) = \sum_t a_{x,t} \bigl( \sum_{i
< t} B_{ti}\, a_{y,i} \bigr)$: at position $t$ the automaton must know
the vector $\sum_{i<t} B_{ti}\, a_{y,i} \in \R^k$, a linear functional
of all digits read so far. It cannot store the digits; it stores their
image under the interface of the current cut. By
\autoref{ex:nonesting}, the interface must be a generating set of the
crossing row module itself:

\begin{lemma}[streaming interfaces]
\label{lem:stream}
For $0 \le t \le n$ fix $V_t \in \R^{r \times t}$ whose rows generate
$\rs(M_t)$ (possible by \autoref{lem:smith}, padding with zero rows;
$V_0$ is empty). Then for every $t \in [n]$ there are $T_t \in \R^{r
\times r}$, $v_t \in \R^{r}$, $R_t \in \R^{k \times r}$ with
\begin{enumerate}
\item $V_t = \bigl[\, T_t V_{t-1} \;\big|\; v_t \,\bigr]$, i.e.\ the
  first $t-1$ columns of $V_t$ equal $T_t V_{t-1}$ and the last column
  is $v_t$;
\item $R_t V_{t-1} = B^{(t)}$, where $B^{(t)} \in \R^{k \times (t-1)}$
  has columns $B_{ti}$ for $i < t$.
\end{enumerate}
\end{lemma}

\begin{proof}
(1) Each row of $M_t$ is indexed by some $(j, l)$ with $j > t$; its
restriction to the first $t-1$ columns is the corresponding row of
$M_{t-1}$, because the same generator $j$ also lies outside $[t-1]$.
Restriction is $\R$-linear, so every row of $V_t$---an $\R$-combination
of rows of $M_t$---restricts to an $\R$-combination of rows of
$M_{t-1}$, i.e.\ into $\rs(M_{t-1}) = \rs(V_{t-1})$; $T_t$ collects the
coefficients, and $v_t$ is the last column. (2) The $l$-th row of
$B^{(t)}$ is the row $(t, l)$ of $M_{t-1}$, because $t \notin [t-1]$;
hence it lies in $\rs(V_{t-1})$.
\end{proof}

The proof is verbatim the field argument, and this is the point: the
restriction calculus of row modules is exact over any commutative ring.
What fails over the ring is only the normalisation of the interface
(\autoref{ex:nonesting}), which the lemma never performs. Merges do not
occur on a chain, so \autoref{ex:ringfail} does not yet bite; it will in
\autoref{sec:master}.

\begin{theorem}[linear layouts]
\label{thm:word}
For all fixed $p, d, k, r$, the class $\lin{p,k,r,d}$ is uniformly
word-automatic. The advice for $(B, \text{layout})$ has length $n + k$
over $p^{d(r^2 + r + kr)} + 1$ advice letters (the presentation's base
alphabet adds the $p^d$ element digits and the padding symbol); the
multiplication automaton has $p^{d(k+r)} + \sum_{j<k} p^{dj} + 1$
states; and every well-formed advice presents a member of the class.
\end{theorem}

\begin{proof}
\emph{Encoding.} An element $(b, a)$ is the word $b_1 \cdots b_k a_1
\cdots a_n$ over the digit alphabet $\{0, \dots, p^d - 1\}$
representing $\R$. The advice is $\mathsf{c}^k \alpha_1 \cdots \alpha_n$
where $\alpha_t$ encodes the triple $(T_t, v_t, R_t)$ of
\autoref{lem:stream} as $r^2 + r + kr$ ring entries.

\emph{The automaton.} States are $(\mathsf{c}, j, \delta)$ with $j < k$,
$\delta \in \R^{j} \times \{0\}^{k-j}$, and $(\mathsf{x}, \delta, w)$
with $\delta \in \R^k$, $w \in \R^r$, plus a sink. On the $j$-th center
position it records $\delta_j := b_{z,j} - b_{x,j} - b_{y,j}$. On
position $t$ it checks $a_{x,t} + a_{y,t} = a_{z,t}$, updates
\[
  \delta \;\leftarrow\; \delta - a_{x,t} \cdot (R_t\, w), \qquad
  w \;\leftarrow\; T_t\, w + v_t \cdot a_{y,t},
\]
and accepts at the end iff $\delta = 0$.

\emph{Correctness.} By induction, after position $t$ the invariant $w =
V_t \, (a_y)|_{[t]}$ holds: it is preserved by \autoref{lem:stream}(1).
Hence at position $t$, $R_t w = R_t V_{t-1} (a_y)|_{[t-1]} = \sum_{i<t}
B_{ti} a_{y,i}$ by \autoref{lem:stream}(2), so the accumulated
subtraction is exactly $C(a_x, a_y)$ of \eqref{eq:cocycle}, and
acceptance means $b_z = b_x + b_y + C(a_x, a_y)$ as required by
\autoref{lem:cocycle}. The universe and equality automata only check the
shape.

\emph{Exactness.} Conversely, an arbitrary well-formed advice defines,
through the same update equations, the correction $\tilde C(a, a') =
\sum_{i < t} \bigl( R_t\, T_{t-1} \cdots T_{i+1}\, v_i \bigr)\, a_t\,
a'_i$, which is $\R$-bilinear and supported on pairs $i < t$; so it
equals \eqref{eq:cocycle} for the form $B_{ti} := R_t T_{t-1} \cdots
T_{i+1} v_i$, and the advice presents $G(B)$ for that form. Its layout
has width $\le r$: the crossing block $M_t$ of the streamed form
factors, by the same telescoping, as $A_t \cdot W_t$ where $W_t \in
\R^{r \times t}$ has columns $T_t \cdots T_{i+1} v_i$, so $\rs(M_t)
\subseteq \rs(W_t)$ is generated by $r$ elements and
$\mrk\bigl(\rs(M_t)\bigr) \le r$ by \autoref{lem:smith}.
\end{proof}

\begin{remark}
The presentation is honest about sizes: the encoding of an element of
$G$ has length $n + k = \log_{p^d} |G|$, and for fixed $\varphi$ the
model-checking run (\autoref{cor:mc}) is linear in $\log|G|$---not in
$|G|$.
\end{remark}

\begin{example}[the clique]
\label{ex:clique}
Let $B_{ji} = 1 \in \R^1$ for all $i < j$: no two generators commute,
and the commutation graph is complete, with pathwidth $n - 1$. Every
crossing block is an all-ones matrix, of module rank $1$; the layout
data can be taken as $T_t = (1)$, $v_t = (1)$, $R_t = (1)$, and the
automaton degenerates to two registers: $w = \sum_{i \le t} a_{y,i}$ and
the deficit $\delta$. Under any vertex-based width measure this is the
hardest instance; under cut-rank it is the easiest.
\end{example}

\begin{example}[extraspecial groups, and valuation-carrying labels]
\label{ex:extraspecial}
Let $n = 2m$, $k = 1$, $d = 1$, and $B_{2t,2t-1} = 1$ for $t \in [m]$: a
perfect matching, presenting the extraspecial-type group of order
$p^{1+2m}$, and \autoref{thm:word} specialises to the uniform
presentation of \autoref{prop:warmA}. Over the ring the same matching
with labels $p^s$, $0 \le s < d$, presents central extensions whose
commutators have order $p^{d-s}$; the module cut-rank of every cut is
still $1$---the width charges a valuation-carrying label fully, as it
must, since the automaton genuinely transports the functional $p^s
a_{y,2t-1}$ across the cut.
\end{example}

\section{The master theorem: distributed centers over the ring}
\label{sec:master}

\autoref{thm:word} fixes the center dimension and streams all central
digits at one end of the layout, so every correction flows upward. This
section removes both restrictions at once: the central generators live on
the layout tree themselves, with commutator targets distributed
arbitrarily, and everything happens over $\R = \ZZ/p^d$.

\subparagraph*{Site trees and cocycle tensors.}
A \emph{site tree} is a binary tree whose $N$ nodes are \emph{sites},
each an $x$-site (generator) or a $z$-site (central generator),
enumerated in post-order; $X, Z \subseteq [N]$ denote the $x$- and
$z$-positions. A \emph{cocycle tensor} over $\R$ is a family $T =
(T_{jiv})$ with $T_{jiv} \in \R$, indexed by $i < j$ in $X$ and $v \in
Z$. It presents $G(T)$: the central extension of $\R^X$ by $\R^Z$ with
correction $C(a, a')_v = \sum_{i<j} T_{jiv}\, a_j a'_i$, to which
\autoref{lem:cocycle} applies verbatim. Elements $(b, a)$ are digit
labellings of the site tree, and a multiplication instance convolves
three of them; the check at a $z$-site $v$ is $(b_z - b_x - b_y)_v =
C(a_x, a_y)_v$, while $x$-sites carry position-wise digit sums.

\begin{definition}[crossing flattenings, width]
\label{def:flattenings}
Let $S$ be a subtree (a post-order interval). Sort the entries $T_{jiv}$
with at least one but not all sites in $S$ into six matrices:
\begin{align*}
F_y &: \text{rows } i \in S, \text{ cols } (j, v) \notin S
  &&\text{($y$-digit functionals, upward)}\\
F_x &: \text{rows } j \in S, \text{ cols } (i, v) \notin S
  &&\text{($x$-digit functionals, upward)}\\
F_m &: \text{rows } (j, i) \subseteq S, \text{ cols } v \notin S
  &&\text{(resolved pair-sums, upward)}\\
F_g &: \text{rows } v \in S, \text{ cols } (j, i) \not\subseteq S
  &&\text{(claims: expected inward flow)}\\
F_{py} &: \text{rows } i \in S, \text{ cols } (j \notin S, v \in S)
  &&\text{(mixed: inside $y$-coefficients of future $x$-digits)}\\
F_{px} &: \text{rows } j \in S, \text{ cols } (i \notin S, v \in S)
  &&\text{(mixed, $x$-side).}
\end{align*}
The \emph{width} of the cut is the maximum of the six module ranks, and
the width of the layout is the maximum over all subtree cuts. For fixed
$p, d, r$ let $\site{p,d,r}$ be the class of pairs (site tree, tensor of
width $\le r$), each presenting $G(T)$.
\end{definition}

Row and column modules of a matrix over $\R$ have the same module rank
(both count the invariant factors below $p^d$ in a Smith normal form),
so the width bounds either side of each flattening. The six matrices
are genuinely different: reshaping a flattening changes its rank. For
the point-target tensor $T_{jiv} = [\, v = \ell(i)\,]$ the flattening
$[(i,v)] \times [j]$ has rank one while $[i] \times [(j,v)]$ has full
rank, and conversely for other tensors; each of the six corresponds to a
distinct kind of traffic a bottom-up evaluator must carry, so the width
must bound each of them, and \autoref{ex:corners}(iv) below shows the
inward kind is genuinely charged.

\begin{example}[the corners, quantitatively]
\label{ex:corners}
(i) With all $z$-sites on a chain above the root, $F_g, F_{py}, F_{px}$
are empty at every layout cut and $\mrk(F_m) \le k$: a form of tree
cut-rank $r$ in the sense of \autoref{sec:prelim} has width $\le \max(r,
k)$. One point deserves a sentence: $F_y$ and $F_x$ split the crossing
block $M_S$ by the orientation of the crossing pair, which is not in
general a rank-bounded operation (zeroing a triangle of an all-ones
matrix blows up its rank). Here the split is a genuine partition of the
\emph{rows} of $M_S$: since $S$ is a post-order interval, an outside
index above it is the larger endpoint of every pair it crosses into $S$,
one below it the smaller, so each row goes wholesale to $F_y$ or to
$F_x$.
(ii) Encode the laminar tree-indexed groups of \autoref{prop:warmB} as
site trees: each inner node $w$ of the shape becomes a chain of two
$x$-sites ($x_w$ directly above $y_w$), each leaf a $z$-site, and
$T_{x_w y_w v} = 1$ for every leaf $v$ below $w$. A subtree cut never
separates a generator pair except along its own chain edge, and every
target lies below its pair, so $F_y = F_x = F_m = 0$; every ancestor
pair owes every central leaf below the cut equally, so each claim
flattening is an all-ones matrix: width $1$.
(iii) \emph{Point targets}: as in (ii) but each pair hits only the
leftmost leaf below it. Width $1$, growing center, non-laminar law: this
family belongs to neither \autoref{thm:word}-style fixed centers nor the
laminar class, so \autoref{thm:master} strictly extends the union of its
corners.
(iv) \emph{Scattered targets}: $m$ private central generators on a chain
below $m$ commuting pairs, pair $t$ targeting $z_t$, have width exactly
$m$: the claim flattening at the $z$-chain is an identity matrix.
\end{example}

\subparagraph*{The protocol.}
The automaton carries six $\R^r$-registers; throughout, ``interface''
means an advice-chosen generating tuple of the indicated module
(\autoref{lem:smith}), and elements of the two bounded modules below are
stored as representatives with respect to such a tuple, canonicalised by
the advice.
\begin{itemize}
\item $w^y = V^y_S\, (a_y)|_{S \cap X}$ and $w^x = V^x_S\, (a_x)|_{S
  \cap X}$, where the rows of $V^y_S$ generate the \emph{column module}
  of $F_y(S)$---each column, indexed by an outside consumer, is a
  functional of the inside digits---and likewise $V^x_S$ for $F_x$;
\item their mixed counterparts $\varphi^y = V^{py}_S (a_y)|_{S \cap X}$
  and $\varphi^x = V^{px}_S (a_x)|_{S \cap X}$ for $F_{py}, F_{px}$:
  inside coefficients that a future outside digit will multiply, the
  product flowing back \emph{into} the subtree's checks;
\item $m$, a representative of the \emph{exports}
  \[
    \mu_S \;:=\; \Bigl( \textstyle\sum_{(j,i) \subseteq S} T_{jiv}\,
    a_{x,j}\, a_{y,i} \Bigr)_{v \notin S} \;\in\; E_S := \rs(F_m(S)),
  \]
  the contribution vector that pairs resolved inside $S$ owe to targets
  outside;
\item $g$, a representative of the \emph{residuals}
  \[
    \rho_S \;:=\; \Bigl( (b_z - b_x - b_y)_v - \textstyle\sum_{(j,i)
    \subseteq S} T_{jiv}\, a_{x,j}\, a_{y,i} \Bigr)_{v \in S \cap Z},
  \]
  which the run requires to lie in the \emph{claim module} $\Gamma_S :=
  \cs(F_g(S))$, rejecting otherwise: the outside can only ever deliver
  elements of $\Gamma_S$, since each future pair contributes an
  $\R$-multiple of its column of $F_g(S)$.
\end{itemize}
$E_S$ and $\Gamma_S$ have module rank $\le r$, so both have at most
$p^{dr}$ elements and representatives in $\R^r$; the state space is
$\R^{6r}$ plus a sink.

Three lemmas supply the transition data. The first is the restriction
calculus, one membership argument instantiated per flattening pair.

\begin{lemma}[restriction calculus]
\label{lem:restrict}
Let $u$ be a node with subtree $S$ and let $C \in \{L, R\}$ index a
child subtree (or the unique child at a unary node). Then:
\begin{enumerate}
\item \emph{folds}: a column of $F_y(S)$, restricted to the rows in $C$,
  is a column of $F_y(C)$; a column of $F_{py}(S)$ restricted to $C$ is
  a column of $F_{py}(C)$ if its target lies in $C$, and a column of
  $F_y(C)$ otherwise; symmetrically on the $x$-side;
\item \emph{read-offs}: for an $x$-site $u$ and $i$ ranging over $C$,
  the coefficient vector $(T_{uiv})_{i \in C}$ is, for each fixed target
  $v$, a column of $F_y(C)$ if $v \notin C$ and a column of $F_{py}(C)$
  if $v \in C$; and for each fixed $i$, the target vector $(T_{uiv})_{v
  \in C \cap Z}$ is a column of $F_g(C)$, while $(T_{uiv})_{v \notin S}$
  is a row of $F_m(S)$;
\item \emph{exports}: restricting a row of $F_m(C)$ to the targets
  outside $S$ gives the corresponding row of $F_m(S)$; hence coordinate
  restriction maps $E_C$ into $E_S$. Restricting it instead to the
  targets in the sibling $C'$ gives a column of $F_g(C')$; hence
  coordinate restriction maps $E_C$ into $\Gamma_{C'}$;
\item \emph{claims}: a column of $F_g(S)$, restricted to the $z$-rows in
  $C$, is a column of $F_g(C)$; hence coordinate restriction maps
  $\Gamma_S$ into $\Gamma_C$.
\end{enumerate}
Consequently all fold and rebase systems below are solvable over $\R$:
whenever a carried tuple must be re-expressed over a target interface,
the relevant vectors lie in the target module.
\end{lemma}

\begin{proof}
Every claim is the same one-line argument. (1): the consumer $(j, v)$ of
a column of $F_y(S)$ satisfies $j, v \notin S \supseteq C$, so the pair
indexes a column of $F_y(C)$; for $F_{py}(S)$ the consumer has $j \notin
S$ and $v \in S$, and restricting to $C$ the target either stays inside
($F_{py}(C)$) or falls outside $C$ ($F_y(C)$). (2): $u \notin C$, so
$(u, v)$ is an admissible column index of the child flattening in each
case, and $(u, i) \subseteq S$ with $v \notin S$ indexes a row of
$F_m(S)$. (3): a pair inside $C$ is inside $S$, and its targets outside
$S$ (resp.\ in $C'$) index the same entries in $F_m(S)$ (resp.\ a column
of $F_g(C')$, since the pair is not inside $C'$). (4): a pair not inside
$S$ is not inside $C$. The final sentence is the definition of the
modules as spans.
\end{proof}

The second lemma is where \autoref{ex:ringfail} lives: the merge of two
sibling interfaces is a well-defined pairing, but not a matrix.

\begin{lemma}[pairing tables]
\label{lem:tables}
Let $u$ be a binary node with children $L, R$, and consider the split
pairs $(i \in L \cap X,\ j \in R \cap X)$, sorted by target into blocks
with $v \notin S_u$, $v \in L$, and $v \in R$. Then the total
contribution of each block,
\[
  \Bigl( \textstyle\sum_{i, j} T_{jiv}\, a_{x,j}\, a_{y,i} \Bigr)_{v},
\]
is a well-defined function of a pair of register values---of $(w^x_R,
w^y_L)$ for the first block, $(w^x_R, \varphi^y_L)$ for the second, and
$(\varphi^x_R, w^y_L)$ for the third---and, as an element-valued map,
lands in $E_{S_u}$, $\Gamma_L$ and $\Gamma_R$ respectively. Each map is
$\R$-bilinear on the images of the two interfaces. In general it does
not extend to a bilinear form on $\R^r \times \R^r$
(\autoref{ex:ringfail}); the advice therefore carries it as a table with
$p^{2rd}$ entries. For $d = 1$ every such map extends to a bilinear
form, and the table may be replaced by $r \times r$ matrices over
$\Fp$.
\end{lemma}

\begin{proof}
Fix a target $v$ of the first block and let $X_v[j, i] = T_{jiv}$. For
fixed $j$, the vector $(T_{jiv})_{i \in L}$ is the column $(j, v)$ of
$F_y(L)$---both $j$ and $v$ lie outside $L$---so every row of $X_v$ lies
in $\rs(V^y_L)$, i.e.\ $X_v = A_v V^y_L$ for some $A_v$. Hence $X_v a =
A_v (V^y_L a)$ depends on $a := (a_y)|_L$ only through $w^y_L$, and
symmetrically---the columns of $X_v$ being columns of $F_x(R)$---the
value $b^{\mathsf T} X_v a$ depends on $b := (a_x)|_R$ only through
$w^x_R$: the contribution is a well-defined function of $(w^x_R,
w^y_L)$, visibly $\R$-bilinear on the images. The target vector of each
split pair, $(T_{ji v})_{v \notin S_u}$, is a row of $F_m(S_u)$, so the
element-valued map lands in $E_{S_u}$. For the second block replace
$F_y(L)$ by $F_{py}(L)$ (the target lies in $L$) and note the target
vectors $(T_{jiv})_{v \in L \cap Z}$ are columns of $F_g(L)$; the third
block is symmetric. Over a field, a bilinear map defined on subspaces
extends to the ambient spaces by completing bases, which yields the
matrices; \autoref{ex:ringfail} shows a map with no such extension over
$\ZZ/4$.
\end{proof}

The third lemma packages the claim discipline; its content is that
elements of $\Gamma_S$ are honest vectors, so the bookkeeping that over
a field is done with coordinates can be done with the elements
themselves.

\begin{lemma}[claim modules]
\label{lem:claims}
$\mrk(\Gamma_S) \le r$ for every subtree $S$, and:
\begin{enumerate}
\item \emph{unique extension}: if $S'$ covers $S$ with $S' \cap Z = (S
  \cap Z) \cup \{v\}$, then an element of $\Gamma_{S'}$ is determined by
  its restriction to $S \cap Z$ together with its coordinate at $v$; and
  at a binary node, an element of $\Gamma_{S_u}$ with $S_u \cap Z = (L
  \cap Z) \cup (R \cap Z)$ is determined by its restrictions to the two
  children. In particular the maps ``extend the claim by a new residual
  coordinate'' and ``join two children's claims'' are partial functions,
  defined exactly on the consistent inputs;
\item \emph{discharge}: subtracting from $g$ any $\R$-combination of
  columns of $F_g(S)$ stays in $\Gamma_S$;
\item \emph{restriction}: by \autoref{lem:restrict}(4), the restriction
  of a claim to a child is a claim of the child, so the extensions and
  joins of (1) are compatible with the invariants below.
\end{enumerate}
At $d = 1$, choosing a full-column-rank submatrix of $F_g(S)$ turns
$\Gamma_S$ into coordinates $\Fp^{\mrk}$ and (1) into the familiar
protocol: commit to the unique coordinates of the residual, reject if
the system is inconsistent, and check that sibling subtrees agree.
\end{lemma}

\begin{proof}
The rank bound is the width; (1) is immediate since elements of
$\Gamma_{S'}$ are vectors indexed by $S' \cap Z$ and the stated data
lists every coordinate; (2) and (3) are closure of a module under
subtraction and \autoref{lem:restrict}(4).
\end{proof}

\begin{theorem}[master theorem: distributed centers over $\ZZ/p^d$]
\label{thm:master}
For all fixed $p$, $d$ and $r$, the class $\site{p,d,r}$ is uniformly
tree-automatic. The multiplication automaton has $p^{6rd} + 1$ states;
the advice expands each site into $O_{p,d,r}(1)$ letters carrying one
ring entry each, over an alphabet of $p^d + O(1)$ letters.
\end{theorem}

\begin{proof}
The advice at each node carries the transition data of
\autoref{lem:restrict}--\autoref{lem:claims} for its subtree: the fold
matrices and injection columns for the four functional interfaces, the
rebase data for $E$ and $\Gamma$ representatives, the read-off data, and
the three pairing tables at binary nodes; element digits repeat along a
site's letter stretch, and the universe automaton enforces the
constancy, so the multiplication automaton may assume it. We maintain,
bottom-up at every subtree $S$, the invariants stated with the
protocol---the four functional register equations, $m$ representing
$\mu_S \in E_S$, $g$ representing $\rho_S \in \Gamma_S$---with the run
alive iff every membership so far held and every digit sum checked.

\emph{Leaves.} At an $x$-site leaf the functional registers initialise
from the interface columns times the site's digits, and $E = \Gamma =
0$. At a $z$-site leaf $v$, the residual $\rho_v = (b_z - b_x - b_y)_v$
must lie in $\Gamma_{\{v\}}$; the letter's table maps it to its
representative or rejects.

\emph{Unary $x$-site $u$ over $C$.} The new pairs are $\{u, i\}$, $i \in
C$. By \autoref{lem:restrict}(2) their coefficient vectors over $i$ lie
in the column modules of $F_y(C)$ or $F_{py}(C)$ according to the
target's position, so the contribution is $a_{x,u} \cdot
(\text{advice-known linear function of } w^y_C \text{ resp.\ }
\varphi^y_C)$; the part with targets outside $S$ enters $m$ (its target
vectors are rows of $F_m(S)$), the part with targets inside discharges
$g$ (its target vectors are columns of $F_g(C)$;
\autoref{lem:claims}(2)). All registers then rebase to the
$S$-interfaces by \autoref{lem:restrict}(1),(3),(4) and absorb the own
digit column.

\emph{Unary $z$-site $v$ over $C$.} The coordinate $v$ leaves the
outside: the letter reads the $v$-coordinate of the represented element
$\mu_C$ (a fixed function of $m$), the residual $\rho_v = (b_z - b_x -
b_y)_v - \mu_C(v)$ arrives, and $g$ extends by
\autoref{lem:claims}(1)---uniquely or rejecting. $m$ rebases along
$E_C \to E_S$ (\autoref{lem:restrict}(3)); the functional interfaces
fold as before.

\emph{Binary node $u$ over $L, R$.} One transition consumes both child
states. The three split-pair blocks are discharged by the pairing
tables of \autoref{lem:tables}: the outside-target block into $m$, the
$L$-target block subtracted from $g_L$, the $R$-target block from
$g_R$. The cross-flows of \autoref{lem:restrict}(3) move the
sibling-target coordinates of $\mu_L$ into a discharge of $g_R$ and
vice versa, and the surviving coordinates of $\mu_L, \mu_R$ rebase into
$E_{S_u}$ and add. The two discharged claims join by
\autoref{lem:claims}(1) into an element of $\Gamma_{S_u}$-so-far,
rejecting on inconsistency; the functional registers fold directly from
the child interfaces to the $S_u$-interfaces by
\autoref{lem:restrict}(1)---the sibling traffic has already been
consumed from the raw child registers, which is why the folds happen at
the merge and not before. Finally the node's own site (an $x$- or
$z$-site) is processed as in the unary cases.

\emph{Root.} Nothing is outside: all six flattenings are empty, $E =
\Gamma = 0$, and the automaton accepts iff the run is alive and $g$
represents $0$. Unwinding the invariants, this holds iff all digit sums
hold and every residual was discharged exactly, i.e.\ $b_z = b_x + b_y +
C(a_x, a_y)$ coordinate-wise: acceptance is $z = x \cdot y$ in $G(T)$.

\emph{Counting.} Six registers with values in $\R^r$ give $p^{6rd}$
live states plus the sink. Each letter carries a bounded number of ring
entries---$\Theta(r^2)$ per fold matrix, $\Theta(r)$ per injection, and
$p^{2rd} \cdot r$ per pairing table (the table value is an
$\R^r$-representative)---so a site expands into $O_{p,d,r}(1)$ letters
of one ring entry each over an alphabet of $p^d$ digit letters plus
finitely many markers.
\end{proof}

\begin{remark}[exactness and the advice set]
\label{rem:exactness}
Unlike in \autoref{thm:word}, not every well-shaped advice presents a
group: a claim rejection can make the streamed relation partial. Take as
the advice set the well-shaped advices whose presented structure
satisfies the first-order group axioms: this set is regular by the
fundamental theorem of automatic structures with
advice~\cite{AbuZaidGraedelReinhardt2017}, every advice in it presents a
class-$2$ group with center of exponent dividing $p^d$, and any tensor
whose cocycle streams through $\R^{6r}$-registers in the above shape has
width $O(r)$, by reading the invariants backwards. The presentation is
therefore exact for the class up to a constant widening.
\end{remark}

\begin{corollary}[string version]
\label{cor:string}
For all fixed $p, d, r$, the subclass of $\site{p,d,r}$ given by
chain-shaped site trees is uniformly word-automatic, with the same state
count.
\end{corollary}

\begin{proof}
On a chain every node has at most one child, so a bottom-up run visits
the sites in a linear order and never branches: the tree automaton
\emph{is} a word automaton reading the advice and the element encodings
leaf-to-root, and the convolution of chain-shaped trees is the
convolution of the corresponding words. Merges---the only genuinely
tree-shaped transitions---do not occur.
\end{proof}

\begin{corollary}[fixed centers, and the warm-ups]
\label{cor:tree}
(i) For all fixed $p, d, k, r$, the class $\tre{p,k,r,d}$ of forms over
$\R$ with a tree layout of module width $\le r$ is uniformly
tree-automatic: place the $k$ central generators on a chain above the
layout root. At every layout cut $F_g, F_{py}, F_{px}$ are empty and
$\mrk(F_m) \le k$, so the site-tree width is $\le \max(r, k)$ by
\autoref{ex:corners}(i), and \autoref{thm:master} applies; on spine
layouts the run simulates the automaton of \autoref{thm:word}.
(ii) The warm-ups are width-one corners: the matching forms of
\autoref{prop:warmA} under (i), and the laminar site trees of
\autoref{ex:corners}(ii), where $\Gamma_S$ is generated by the all-ones
vector, the represented claim is the single deficit owed to every leaf
below, the join of \autoref{lem:claims}(1) is ``sibling subtrees must
owe equally'', and the protocol collapses to the $(p+1)$-state automaton
of \autoref{prop:warmB}. By \autoref{ex:corners}(iii) the master theorem
is strictly stronger than the union of its corners.
\end{corollary}

\section{The wealth of the class}
\label{sec:span}

This section maps what bounded width does and does not buy. The
parameter interfaces exactly with graph rank-width
(\autoref{prop:rankwidth}); the class is closed under central and direct
products, with the width of a product being the \emph{maximum}, not the
sum, of the widths of its factors (\autoref{prop:products}); the
classical width measures of the commutation graph bound it
(\autoref{prop:widths}); and at the tame end it collapses, while
generically it explodes (\autoref{prop:collapse},
\autoref{prop:generic}). \autoref{tab:constructions} places the named
constructions.

\begin{proposition}[rank-width interface]
\label{prop:rankwidth}
Extend tree layouts by \emph{structural nodes} carrying no generator
(the element encodings carry a forced $0$ digit there; the automata are
unchanged). Then for $k = 1$ and $d = 1$: a form belongs to
$\tre{p,1,r,1}$ (with structural nodes) if and only if the
$\Fp$-labelled commutation graph of $B$ has rank-width at most $r$ over
$\Fp$, and a rank-decomposition converts to a layout without changing
the width.
\end{proposition}

\begin{proof}
A rank-decomposition is a subcubic tree with the vertices at its leaves;
rooting it at an arbitrary edge yields a binary tree whose internal
nodes become structural. The generator set of any subtree is one side of
an edge cut of the decomposition, and its crossing block is the
bipartite adjacency block whose rank defines cut-rank. Conversely a
layout converts to a rank-decomposition by replacing every generator at
an internal node with a structural node whose new leaf child carries the
generator: this adds only singleton cuts, of cut-rank $\le 1$, so the
width is unchanged for $r \ge 1$.
\end{proof}

\begin{proposition}[closure under central and direct products]
\label{prop:products}
Let $T_1, T_2$ be cocycle tensors over $\R$ on disjoint site trees, of
widths $r_1, r_2$. Grafting the two trees onto a common root presents
$G(T_1) \times G(T_2)$ (a direct product; identifying central generators
instead gives central products) with width $\max(r_1, r_2, 1)$. The
same holds for any finite number of factors.
\end{proposition}

\begin{proof}
The combined tensor is block-diagonal: no commutation pair and no
target crosses between the blocks. A subtree cut of the grafted layout
either lies inside one block---where every flattening equals that
block's flattening padded by zero rows and columns, of unchanged module
rank---or contains one block entirely, in which case the entries
crossing the cut are exactly the entries crossing the induced cut of the
other block. Identifying central generators merges $z$-sites; the claim
flattenings gain summed rows, which cannot raise the module rank beyond
the maximum plus the all-ones pattern of rank $1$.
\end{proof}

The proposition is the model-theoretic heart of the section: the
operations that preserve uniform automaticity---central products, direct
products, direct factors with abelian groups, raising the center's
exponent---\emph{are} group-theoretic constructions, and the advice of a
product is the concatenation of the factors' advices under a common
root. One compiled formula therefore serves every iterated combination
of every family in \autoref{tab:constructions} at once: constructions
classically studied in isolation become a single decidable family.

\begin{table}[t]
\centering\small
\renewcommand{\arraystretch}{1.15}
\begin{tabular}{@{}p{4.4cm}p{5.6cm}p{3.2cm}@{}}
\toprule
Construction & Setting & Width \\
\midrule
finite abelian $p$-group & no commutation & $0$ \\
extraspecial $p^{1+2m}$ & matching, $k = 1$ & $1$ \\
Heisenberg over $\Fp^{n}$ & symplectic normal form & $1$ \\
$n$-qubit Pauli group & $p = 2$, clique & $1$ \\
special $p$-groups & fixed $k$, form of width $\le r$ & $r$ \\
shared-center graph group & $[x_i, x_j] = y$ on edges & graph rank-width \\
tree-indexed laminar & distributed center, laminar & $1$ \\
point-target family & distributed, non-laminar & $1$ \\
exponent-$p^d$ centers & any of the above over $\R$ & unchanged \\
central / direct products & block-diagonal tensors & $\max$ of blocks \\
\midrule
ultraspecial (Heisenberg over $\mathbb{F}_{p^m}$) & field-multiplication
tensor & $\Theta(m)$ \\
Mekler groups & private target per non-edge & $\Omega(n)$ \\
relatively free class-$2$, $\exp p$ & fully free & not
automatic~\cite{NiesStephan2023} \\
\bottomrule
\end{tabular}
\caption{Named constructions as layouts. The rows below the line are
excluded---and are exactly the model-theoretically wild objects.}
\label{tab:constructions}
\end{table}

\begin{proposition}[classical width measures]
\label{prop:widths}
Let $\Gamma$ be the commutation graph of a form $B$ over $\R$ (edges =
non-commuting pairs).
\begin{enumerate}
\item If $\Gamma$ has a vertex cover $C$ of size $c$, then every layout
  that places $C$ first has width $\le c$; in particular $\lcr(B) \le
  c$. An arbitrary layout has width $\le kc$, and the factor $k$ is
  necessary.
\item If $\Gamma$ has pathwidth $w$, then $\lcr(B) \le w + 1$.
\item If $\Gamma$ has treewidth $w$, then $\tcr(B) \le k(w+1)$.
\end{enumerate}
\end{proposition}

\begin{proof}
All three are counting arguments---the module rank of a matrix is at
most its number of nonzero rows and at most its number of nonzero
columns---and hold over $\R$ verbatim. (1) Put the cover first. A cut $S
\subseteq C$ offers at most $|S| \le c$ columns. A cut $S \supseteq C$
contains every edge's cover endpoint, so the nonzero columns of the
crossing block are indexed by $C$: again module rank $\le c$. In an
arbitrary layout the nonzero entries lie in the $c_1 \le |C \cap S|$
columns of the inside cover vertices and the $\le k c_2$ rows of the
outside ones, whence module rank $\le c_1 + k c_2 \le kc$; a star with
$c = 1$ whose edge to the $j$-th leaf is labelled $e_j$ attains $k$ at
the cut separating the center from its leaves. (2) Order the vertices by
a path decomposition; the earlier endpoints of crossing edges at any cut
are active vertices, so nonzero columns number $\le w + 1$. (3) Expand
each bag of a tree decomposition into a chain introducing its new
vertices, and take this as the layout. The bag of the current edge of
the decomposition separates the subtree from the rest, so every
crossing pair has an endpoint in it: nonzero columns for
below-endpoints ($\le w+1$) and $k$ rows per above-endpoint bound the
module rank by $k(w+1)$.
\end{proof}

\begin{proposition}[collapse at the tame end]
\label{prop:collapse}
Let $p$ be odd and $d = 1$.
\begin{enumerate}
\item If $\Gamma$ has a vertex cover of size $c$, then $G(B) \cong H
  \times \Zp^{\,m}$ with $|H| \le p^{\,c(k+1) + k}$; for fixed $c$ and
  $k$ the class of such groups is, up to isomorphism, a finite list of
  cores times elementary abelian factors.
\item For $k = 1$, every $G(B)$ is isomorphic to (extraspecial of
  exponent $p$) $\times$ (elementary abelian): the symplectic normal
  form aligns every form with a matching.
\end{enumerate}
\end{proposition}

\begin{proof}
(1) The radical $\bigcap_{l \le k} \operatorname{rad}(B_l)$ has
codimension $\le c(k+1)$: pairing against $e_j$ for $j$ outside the
cover constrains only the $c$ cover coordinates of a vector (every edge
at $j$ ends in the cover), and pairing against the $c$ cover vertices
adds at most $ck$ further linear conditions. Split off the radical (for
odd $p$ its preimage is elementary abelian and complements off the
commutator support), so $|H| \le p^{\,c(k+1)+k}$; the bound's shape is
right: a star with $c = 1$, $k = n - 1$ and $B_{ui} = e_i$ has trivial
radical and $|H| = p^{2n-1}$. (2) is the classical normal form of an
alternating form: rank $2m$, matching of $m$ hyperbolic pairs, radical
split off.
\end{proof}

The reading for the mixed audience: at $k = 1$, $d = 1$ (and bounded
vertex cover for any $k$) the parameter buys presentation-theoretic
rather than isomorphism-theoretic generality---the clique of
\autoref{ex:clique} is abstractly a Heisenberg group times an abelian
group, but the automaton multiplies in the \emph{given} presentation
without computing the normal form. The isomorphism-theoretic content of
the class lives at $k \ge 2$, where pairs of alternating forms already
have infinitely many normal-form families and triples are
wild~\cite{GrochowQiao2021}; at $d \ge 2$, where the center's exponent
is an invariant the field case cannot see; and, already at width one,
with the distributed centers of \autoref{thm:master}, whose laminar and
point-target families have commutator subgroups of rank $\Theta(n)$ and
so escape the collapse of \autoref{prop:collapse} entirely.

\begin{proposition}[generic forms are wide]
\label{prop:generic}
Let $k = 1$, $d = 1$, and let $B$ be uniformly random. With high
probability, every linear layout of $B$ has width $\ge n/2 -
O(\sqrt{n})$ and every tree layout has width $\ge n/3 - O(\sqrt{n})$;
in particular $\tcr(B) = \Omega(n)$.
\end{proposition}

\begin{proof}[Proof sketch]
For a fixed set $S$ with $|S| = m \le n/2$, the crossing block is a
uniform random $(n-m) \times m$ matrix over $\Fp$, and $\Pr[\rk < m -
s] \le p^{-s(n-2m+s)} \le p^{-s^2}$. A union bound over all $\le 2^n$
sets with $s = c\sqrt{n}$ leaves all sets of size $m \in [n/3, n/2]$
with rank $\ge m - c\sqrt n$. Every linear layout has a cut of size
$\lfloor n/2 \rfloor$, and every binary tree has a subtree whose
interval size lies in $[n/3, 2n/3]$; complement if necessary.
\end{proof}

\subparagraph*{The dividing line.}
The excluded rows of \autoref{tab:constructions} are excluded provably,
and they are exactly the model-theoretically wild objects.

\begin{example}[ultraspecial groups]
\label{ex:ultraspecial}
The Heisenberg group over the field $\mathbb{F}_{p^m}$---an
\emph{ultraspecial} group---has generators $a_1, \dots, a_m, b_1,
\dots, b_m$ and commutation tensor the multiplication of
$\mathbb{F}_{p^m}$, written over $\Fp$ with $k = m$: $[a_i, b_j]$ is
the field product $e_i e_j$ in the center. Its width is $\Theta(m)$
under every layout: take a cut $S$ with $2m/3 \le |S| \le 4m/3$, let
(by symmetry) the $a$-side be the majority side of $S$, and pick any
$b_j$ outside $S$---one exists since $S$ is not everything. The columns
of the crossing block indexed by the $a_i \in S$ contain the vectors
$(e_i e_j)_l$, and a vanishing combination forces $(\sum_i c_i e_i)\,
e_j = 0$, hence $\sum_i c_i e_i = 0$ since $e_j$ is invertible: those
columns are linearly independent, and the rank is $\ge |S|/2 - 1 =
\Omega(m)$.
\end{example}

\begin{example}[Mekler groups]
\label{ex:mekler}
Mekler's construction~\cite{Mekler1981} interprets a graph in a
class-$2$ exponent-$p$ group whose commutator structure gives each
non-adjacent generator pair a \emph{private} central target. Private
targets defeat every layout: fix any site tree and a subtree $S$
containing between one and two thirds of the $x$-sites. Each pair
$\{u, v\}$ split by $S$ with a private target $z_{uv}$ contributes,
depending on where $z_{uv}$ lies, a row of $F_y/F_x$ or of $F_g$ whose
support---the private column $(v, z_{uv})$, resp.\ the private pair
column---is disjoint from every other such row's support. Disjointly
supported nonzero rows are linearly independent, so the width is at
least the number of inside endpoints with a split private pair, which
is $\Omega(n)$ whenever the non-adjacency is dense---as it is for the
graphs Mekler's interpretation uses.
\end{example}

Together with \autoref{prop:generic} and the theorem of Nies and
Stephan that the relatively free class-$2$ exponent-$p$ group is not
word-automatic~\cite{NiesStephan2023}, the picture is consistent in
both directions: bounded rank-width is a thin, structured slice, and
the constructions that certify wildness---generic forms, field
multiplication, private targets, full freeness---all have width
$\Omega(n)$. Within class-$2$ groups with center of exponent $p^d$, the
class of \autoref{thm:master} is the tame complement of Mekler's
construction.

\section{Algorithmic consequences}
\label{sec:consequences}

The generic machinery of uniformly automatic classes
\cite{AbuZaidGraedelReinhardt2017} yields, for the master class and
hence for every subclass of \autoref{sec:span}:

\begin{corollary}[compiled model checking]
\label{cor:mc}
For every first-order formula $\varphi(x_1, \dots, x_m)$ in the
signature of groups there is (effectively) an automaton
$\mathfrak{A}_\varphi$ such that for every advice $\adv$ in the advice
set and all elements $u_1, \dots, u_m$ of $G_\adv$:
\[
  G_\adv \models \varphi(\bar u) \iff \mathfrak{A}_\varphi \text{
  accepts } \adv \conv u_1 \conv \dots \conv u_m .
\]
In particular, once $\varphi$ is compiled, deciding it on a member of
the class takes time $O(N) = O(\log |G|)$.
\end{corollary}

\begin{corollary}[decidable uniform theory]
\label{cor:theory}
The advice sets of \autoref{thm:word} and (by \autoref{rem:exactness})
of \autoref{thm:master} are regular. Hence the uniform first-order
theory $\{\varphi : G_\adv \models \varphi \text{ for all } \adv\}$ is
decidable, and so are existential class queries (``is there a member
satisfying $\varphi$?''). The same holds for extensions of first-order
logic by regularity-preserving quantifiers, e.g.\ cardinality and
Ramsey quantifiers~\cite{AbuZaidGraedelReinhardt2017}.
\end{corollary}

\begin{example}
The center is definable, $\mathrm{Cen}(x) \equiv \forall u \forall v
\forall w\, (M(x,u,v) \wedge M(u,x,w) \to v = w)$, so \autoref{cor:mc}
decides centrality of a given element in time $O(\log|G|)$ on every
member; likewise commutativity of the member, existence of central
elements of order $4$ at $p = 2$ (distinguishing the Pauli triangle
from the dihedral matching), the exponent of the center for $d > 1$, or
any fixed isoclinism-flavoured first-order invariant. One automaton per
property serves the entire class---and, by \autoref{prop:products},
every central and direct product of its members.
\end{example}

\section{Implementation, machine verification, and feasibility}
\label{sec:implementation}

The constructions of this paper are implemented in the open-source
library \AutStr\AutStrCite{}: \texttt{CutRankGroups}
(\autoref{thm:word}), \texttt{CutRankTreeGroups}
(\autoref{cor:tree}(i)) and \texttt{CocycleRankWidthGroups}
(\autoref{thm:master} at width $1$), each parameterised by the depth
$d$ and equipped with a layout compiler that realises the lemmas of
\autoref{sec:word} and \autoref{sec:master} by linear algebra over
$\R$. The implementation serves two purposes: it \emph{machine-verifies}
the constructions, and it measures where the algorithmic meta-theorems
of \autoref{sec:consequences} are practical.

\subparagraph*{Machine verification.}
The compilers are self-certifying: every advice constant is obtained by
solving a linear system whose solvability is an instance of
\autoref{lem:stream}, \autoref{lem:restrict}, \autoref{lem:tables} or
\autoref{lem:claims}, guarded by an assertion---a compilation that
succeeds is a machine-checked certificate that the memberships hold on
that instance. The multiplication automata are checked against the
group law of \autoref{lem:cocycle} exhaustively on structured layouts
(spines and balanced trees; zero, complete, matching and star forms;
the laminar, point-target and mixed-claim site trees) and on thousands
of randomly generated width-admissible forms and tensors over $\ZZ/4$,
$\ZZ/8$ and $\ZZ/9$, including instances realising
\autoref{ex:ringfail} and \autoref{ex:nonesting}; no assertion has
fired and no mismatch occurred on any width-admissible input. At
$d = 1$ the fixed-center ring code paths reproduce the field
constructions byte-for-byte. The distributed-center implementation
realises the protocol of \autoref{thm:master} exactly---six registers,
one-step merges, table-driven letters---at any width $r$ and depth $d$;
the advice of the fixed-center classes can be emitted either as one
letter per position or in the streamed one-entry-per-letter form, which
is what makes widths $r \ge 2$ over the ring representable in
practice.

\subparagraph*{Feasibility.}
%% TODO: refresh all numbers in this subparagraph from the extended
%% benchmark run (10-minute per-task timeout, <= 10 h budget) before
%% submission; the qualitative claims below are from the first sweep.
Two evaluation paths are available. The \emph{compiled} path builds the
presentation automata explicitly and evaluates formulas by automata
products and projections---the promise of \autoref{cor:mc}, with the
one-time compilation cost amortised over members. The \emph{implicit}
path never builds an automaton: it evaluates a formula directly over
the transition functions, taking products of composite states on the
fly and handling quantifiers by an on-the-fly powerset construction
over the element digits. A sweep on a cloud VM (16 cores, 64\,GB,
per-task memory caps) locates the practical boundary of each path:
\begin{itemize}
\item the compiled path builds the base automata of the field classes
  and of small ring members (word members over $\ZZ/4$; with streamed
  letters also width-$2$ word members over $\ZZ/4$, in under a minute),
  and compiles typical fixed formulas over them in seconds;
\item the base automata of larger ring members ($\ZZ/8$ and $\ZZ/9$
  word members, $\ZZ/4$ tree members) and of the distributed-center
  protocol are out of reach of explicit construction---the
  multiplication automaton is a product over the advice alphabet---but
  the \emph{implicit} path decides first-order formulas on all of them,
  agreeing with the compiled path on every instance where both
  complete, with no disagreement observed anywhere in the sweep;
\item the limiting resource of the implicit path is quantifier
  alternation, not memory: no out-of-memory event occurred in the
  sweep; deeply alternating sentences over the largest ring members
  time out instead, as the on-the-fly powersets grow.
\end{itemize}
The practical reading: for a fixed query answered on many members, the
compiled path is preferable whenever the member's automata build at
all, and the crossover is quick since a compiled decision is a single
linear pass; the implicit path extends every meta-theorem of
\autoref{sec:consequences}, as a decision procedure if not as a
compiled linear-time algorithm, to members far beyond the reach of
compilation.

\section{Open problems}
\label{sec:open}

\begin{enumerate}
\item \emph{Converse / characterisation.} Does uniform (word-, tree-)
  automaticity of a class of forms or tensors over $\R$ force bounded
  width under suitably chosen layouts? A positive answer would be a
  Courcelle--Oum-style characterisation~\cite{CourcelleOum2007} of
  automaticity on the class-$2$ stratum. \autoref{prop:generic}, the
  negative result of~\cite{NiesStephan2023}, the width-$\Omega$ bounds
  of \autoref{ex:ultraspecial} and \autoref{ex:mekler}, and the
  exactness direction of \autoref{rem:exactness}---bounded-interface
  streams have bounded width---are the supporting evidence;
  \autoref{ex:corners}(iv) shows the six flattenings charge exactly the
  traffic a bottom-up automaton must carry.
\item \emph{Isometry in P.} Is isometry of alternating matrix spaces of
  bounded $\Fp$ rank-width decidable in polynomial time? This is the
  matrix-space analogue of isomorphism testing for graphs of bounded
  rank-width~\cite{GroheSchweitzer2015} and would identify a tame
  island inside a tensor-isomorphism-complete
  problem~\cite{GrochowQiao2021}; with distributed centers and $d > 1$
  the same question extends to isomorphism of the width-$r$ tensor
  classes of \autoref{thm:master}. Computing a witnessing layout is
  itself nontrivial; for $k = 1$, $d = 1$ rank-decompositions are
  FPT-computable~\cite{HlinenyOum2008}.
\item \emph{Compiled evaluation over the general protocol.} The
  micro-operations of the distributed-center automaton act totally on
  its register space, so its flat transition relation is dense, and
  compiling quantified formulas over it currently exceeds commodity
  memory (\autoref{sec:implementation}); the implicit path covers the
  gap as a decision procedure. Closing it in the compiled
  path---evaluating first-order queries over the protocol automaton
  within practical resources---is an engineering problem with a
  mathematical flavour: the transition structure is linear algebra over
  $\R$, and a representation exploiting that structure should exist.
\item \emph{Beyond exponent $p^d$ and class $2$.} Arbitrary bounded
  abelian quotients add a carry layer to the extension data; class $3$
  makes the correction a degree-three stream. Does the cut-rank
  philosophy---bound the module rank of every flattening a one-pass
  evaluation must transport---extend along the Lazard correspondence to
  higher class, and where exactly does it stop short of the
  non-automaticity of~\cite{NiesStephan2023}?
\end{enumerate}

\section*{Declaration of Generative AI Usage}
\label{sec:ai-disclosure}

In accordance with the Schloss Dagstuhl Publishing policy, the authors declare the use of generative artificial intelligence (AI) in the preparation of this manuscript.

Specifically, large language models (\texttt{Claude Fable 5} and \texttt{Claude Opus}) were utilized in interactive sessions to assist in drafting the narrative text of the paper. This drafting process was strictly guided by detailed instructions, mathematical directions, and structural proof and construction sketches provided entirely by the human authors. Conversely, all literature research, citation collection, and the final bibliography reflect exclusively the authors' own independent scholarly research without AI intervention.

The authors have reviewed, verified, and edited all generated outputs and take full, final responsibility for the correctness, scientific integrity, and overall content of this work.


\bibliography{refs}

\end{document}
